---
title: "GM DLP Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(RColorBrewer)

Sys.setenv("DISPLAY"=":0.0")
```

## Read copy number data    

```{r readmetrics}
metrics = read_csv('/juno/work/shah/tantalus/SC-3423/results/annotation/A73044A_metrics.csv.gz')
metrics %>% head()
```


```{r}
p <- metrics %>%
  ggplot(mapping = aes(x=quality, y=total_reads, color=is_s_phase)) + 
  geom_point() +
  theme(legend.position = "left")
ggMarginal(p, type="histogram", groupColour=TRUE, groupFill=TRUE)
```



```{r readdata}
cn = read_csv(
  '/juno/work/shah/tantalus/SC-3423/results/hmmcopy/A73044A_reads.csv.gz',
  col_types = cols(
    chr = col_factor(NULL),
    cell_id = col_factor(NULL)
  ))
```


```{r readdata}
cn %>% head()
```

```{r}
cn %>% filter(cell_id == "SA928-A73044A-R08-C59", gc > 0) %>%
  mutate(chr = factor(chr, levels = c(1:22, "X", "Y"))) %>%
  ggplot() +
  geom_point(mapping = aes(x = start, y = reads), size = 0.5) +
  facet_grid(~chr, scales = "free_x", space="free_x", switch = "x") +
  theme(panel.spacing = unit(0.05, "cm")) +
  scale_x_continuous(breaks = c()) +
  xlab('chromosome')
```


```{r}
cn %>% filter(cell_id == "SA928-A73044A-R08-C59", gc > 0) %>%
  ggplot(aes(x = gc, y = reads)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "loess") +
  theme_bw()
```


```{r}
cn %>%
  dplyr::filter(cell_id == "SA928-A73044A-R08-C59", gc > 0) %>%
  mutate(cn_state = factor(state, names(cn.colors))) %>%
  mutate(chr = factor(chr, levels = c(1:22, "X", "Y"))) %>%
  ggplot() +
  geom_point(mapping = aes(x = start, y = copy, colour = cn_state), size = 0.5) +
  facet_grid(~chr, scales = "free_x", space="free_x", switch = "x") +
  theme(panel.spacing = unit(0.05, "cm")) +
  scale_x_continuous(breaks = c()) +
  scale_color_manual(values = cn.colors, labels = names(cn.colors), drop = FALSE) +
  scale_y_continuous(breaks = seq(0, 6, by = 1), limit = c(0, 6)) +
  xlab('chromosome')
```


```{r}
cn %>% filter(cell_id == "SA928-A73044A-R08-C58", gc > 0) %>%
  ggplot(aes(x = gc, y = reads)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "loess") +
  theme_bw()
```


```{r}
cn %>%
  filter(cell_id == "SA928-A73044A-R08-C58", gc > 0) %>%
  mutate(cn_state = factor(state, names(cn.colors))) %>%
  mutate(chr = factor(chr, levels = c(1:22, "X", "Y"))) %>%
  ggplot() +
  geom_point(mapping = aes(x = start, y = copy, colour = cn_state), size = 0.5) +
  facet_grid(~chr, scales = "free_x", space="free_x", switch = "x") +
  theme(panel.spacing = unit(0.05, "cm")) +
  scale_x_continuous(breaks = c()) +
  scale_color_manual(values = cn.colors, labels = names(cn.colors), drop = FALSE) +
  scale_y_continuous(breaks = seq(0, 8, by = 1), limit = c(0, 8)) +
  xlab('chromosome')
```

## Plot a single cell

Select a cell then plot normalized binned reads across the genome.

```{r plotsingle, echo=FALSE}
cell = filter(cn, cell_id == "SA922-A90554B-R35-C19")

cell$chr <- factor(cell$chr, levels = c(1:22, "X", "Y"))

cn.colors <- c(rev(brewer.pal(n = 3, "Blues"))[1:2], "#CCCCCC", tail(brewer.pal(n = 8, "OrRd"), 6))
cn.colors <- c(cn.colors, cn.colors[c(9, 9, 9, 9)])
names(cn.colors) <- 0:12

cell$cn_state = factor(cell$state, names(cn.colors))

ggplot(data = cell) +
  geom_point(mapping = aes(x = start, y = copy, colour = cn_state), size = 0.5) +
  facet_grid(~chr, scales = "free_x", space="free_x", switch = "x") +
  theme(panel.spacing = unit(0.05, "cm")) +
  scale_x_continuous(breaks = c()) +
  scale_color_manual(values = cn.colors, labels = names(cn.colors), drop = FALSE) +
  scale_y_continuous(breaks = seq(0, 18, by = 2), limit = c(0, 12)) +
  xlab('chromosome')
```

## Exploring cells with different properties

Lets make a function out of our cell plotting code and use it to plot cells with different properties.

```{r plotfunc, echo=FALSE}
plot_cell <- function(cn, plot_cell_id) {
  cell = filter(cn, cell_id == plot_cell_id)
  
  cell$chr <- factor(cell$chr, levels = c(1:22, "X", "Y"))

  cn.colors <- c(rev(brewer.pal(n = 3, "Blues"))[1:2], "#CCCCCC", tail(brewer.pal(n = 8, "OrRd"), 6))
  cn.colors <- c(cn.colors, cn.colors[c(9, 9, 9, 9)])
  names(cn.colors) <- 0:12

  cell$cn_state = factor(cell$state, names(cn.colors))

  ggplot(data = cell) +
    geom_point(mapping = aes(x = start, y = copy, colour = cn_state), size = 0.5) +
    facet_grid(~chr, scales = "free_x", space="free_x", switch = "x") +
    theme(panel.spacing = unit(0.05, "cm")) +
    scale_x_continuous(breaks = c()) +
    scale_color_manual(values = cn.colors, labels = names(cn.colors), drop = FALSE) +
    scale_y_continuous(breaks = seq(0, 18, by = 2), limit = c(0, 12)) +
    xlab('chromosome')
}
```

A cell with very low coverage:

```{r plotcell1, echo=FALSE}
plot_cell(cn, "SA922-A90554B-R23-C39")
```
A low quality cell:

```{r plotcell1, echo=FALSE}
plot_cell(cn, "SA922-A90554B-R23-C03")
```

A cell in s-phase:

```{r plotcell1, echo=FALSE}
plot_cell(cn, "SA922-A90554B-R23-C64")
```
Note that 

```{r plotcell1, echo=FALSE}
cn %>% filter(cell_id == "SA922-A90554B-R27-C22", gc > 0) %>%
  ggplot() +
  geom_point(mapping = aes(x = gc, y = reads, colour = state), size = 0.5)
```

A cell with catastophic mitotic error including complete loss of 9q.

```{r plotcell1, echo=FALSE}
plot_cell(cn, "SA922-A90554B-R23-C38")
```
A cell in s-phase.

```{r plotcell1, echo=FALSE}
plot_cell(cn, "SA922-A90554B-R23-C64")
```


low coverage: SA922-A90554B-R23-C39

low coverage: SA922-A90554B-R23-C03

low coverage: SA922-A90554B-R23-C03


```{r}
cell = filter(cn, cell_id == "SA922-A90554B-R35-C19")

  cell = filter(cn, cell_id == plot_cell_id)
  
  cell$chr <- factor(cell$chr, levels = c(1:22, "X", "Y"))

  cn.colors <- c(rev(brewer.pal(n = 3, "Blues"))[1:2], "#CCCCCC", tail(brewer.pal(n = 8, "OrRd"), 6))
  cn.colors <- c(cn.colors, cn.colors[c(9, 9, 9, 9)])
  names(cn.colors) <- 0:12

  cell$cn_state = factor(cell$state, names(cn.colors))

cn %>% filter(cell_id == "SA922-A90554B-R35-C19") %>%
  mutate()
ggplot(data = cell) +
  geom_rect(mapping = aes(xmin = start, xmax=end, ymin = copy - 0.2, ymax=copy + 0.2), color="black", fill="white") +
  facet_grid(~chr, scales = "free_x", space="free_x", switch = "x") +
  theme(panel.spacing = unit(0.05, "cm")) +
  scale_x_continuous(breaks = c()) +
  scale_y_continuous(breaks = seq(0, 18, by = 2), limit = c(0, 12)) +
  xlab('chromosome')
```


```{r}

plot_heatmap <- function(gr){
  #' plot a heatmap with chromosome boundaries
  #' the order of the rows can be customized here
  #' its a simple distance based clustering
  
  mat <- as.matrix(GenomicRanges::mcols(gr))
  mat <- t(mat)
  hr <- hclust(dist(mat), method = "average")
  hr = as.dendrogram(hr)
  
  # additional chromosome annotations
  nbins <- chrom_bdry[length(chrom_bdry)]
  abline_x = chrom_bdry[-1]/nbins
  x_label <- names(nbins_aggr)
  mid.point <- (nbins_aggr/2)/nbins
  mid.point <- abline_x - mid.point
  abline_ids <- rep(1:length(abline_x),each=2)
  abline_y <- rep(c(0,1), times=length(abline_x))
  rownames(abline_x) <- NULL
  abline_x <- rep(abline_x,each=2)
  rm(nbins)
  
  #annotation to label chromosomes
  ha_column = ComplexHeatmap::HeatmapAnnotation(cn = function(index) {
    grid.text(x_label,x=mid.point,y=1,just = c("center", "top"),gp=gpar(col="#202020",fontsize=6))
  })
  
  #the main heatmap
  mat[mat>=6] = ">=6"
  colors = structure(
    rev(c("#3783bb","#a7d0e4","#f7f6f6","#fca082","#fb694a","#e32f27","#b11218")),
    names = rev(c("0", "1", "2", "3","4","5",">=6")))
  
  hm = ComplexHeatmap::Heatmap(matrix = mat,
    name = "ploidy",
    col = colors,
    cluster_rows = hr,
    cluster_columns = FALSE,
    show_row_names = FALSE,
    bottom_annotation = ha_column,
    column_title = "CNV heatmap @ 2Mb resolution")
  
  ComplexHeatmap::draw(hm, row_dend_side = "left")
  ComplexHeatmap::decorate_heatmap_body("ploidy", {
    grid.polyline(
      x= abline_x,
      y = abline_y,
      id = abline_ids,
      gp = gpar(lty = 1, lwd = 1.5))
  }) 
}

```


```{r}
cn_gr = cn[, c("chr", "start", "end", "state", "cell_id")] %>%
  pivot_wider(names_from = cell_id, values_from = state) %>%
  GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns=TRUE,ignore.strand=TRUE)
```

```{r}
cn_gr %>%
  plot_heatmap()
```


